Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.

### 2022-07-25 5_artigos/statistic-in-medicine/draft/DRAFT2/draft-sm-lineu.tex
atualiza rascunhos e dissertacao

\title{Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models}

@@

\author[2]{Wagner Hugo Bonat}

\author[3]{Ligia de Oliveira Carlos}

\author[3]{Antônio Carlos Ligocki Campos}

@@

\address[2]{\orgdiv{Department of Statistics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[3]{\orgdiv{Department of Surgical Clinic}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

@@

Clinical trials are common in medical research. This type of study usually generates data with multiple non-Gaussian responses and time-dependent observations. The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. Our proposal is to use the Wald test to perform general hypothesis tests, ANOVAs, MANOVAs and multiple comparison tests on regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of  correlation structures. The test was verified based on simulation studies and the results showed that the further the hypothesis tested is from the true values of the parameters, the greater the percentage of rejection of the null hypothesis. The lowest rejection percentages were observed when the hypothesis tested corresponded to the real values of the parameters. Furthermore, the larger the sample size, the greater the power of the test. The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

@@

\jnlcitation{\cname{%
\author{Freitas LAC}, 
\author{Bonat WH}, 
\author{Carlos LO}, and 
\author{Campos. ACL}} (\cyear{2022}), 
\ctitle{Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models}, \cjournal{Statistics in Medicine}}

@@

%\subsection{Estudos clínicos}

Clinical trials represent one of the most common procedures used to assess whether a new treatment is safe, effective or even better than a standard treatment. These are procedures widely used in medical research with the aim of evaluating interventions on a group of individuals. In general, the data generated by clinical trials are not trivial and, for analysis, there is a need for statistical methods capable of dealing with all the specifics of this type of study.

According to \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

%\subsection{Problemas na área médica que envolvem múltiplas respostas}

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); \cite{schmitz2019effect} evalueated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; \cite{duma2019characterization} tested the association between certain exclusion criteria and trial characteristics. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

%\subsection{Introdução ao Dataset}

Similarly to the aforementioned articles, we are interested in the analysis of multiple outcomes in the context of longitudinal data analysis. The study we shall describe was conducted with the objective of evaluating the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. This is a problem with two response variables: a score that characterizes compulsion and the number of symptoms presented that characterize addiction.

In this study, a set of subjects was divided into two groups: one of them received a placebo and the other received the treatment. In addition, individuals were evaluated at 3 different times: the first evaluation was performed approximately 10 days before surgery; follow-up assessments were performed approximately three months and one year postoperatively.

%\subsection{McGLM}

It is important to note that this is a multivariate problem (contains two response variables), with non-Gaussian responses and that the observations that make up the data set cannot be considered independent, considering that measurements taken on the same individual at different times tend to be correlated. Therefore, it is a problem that traditional modeling techniques would be difficult to apply. It is necessary to use some methodology that supports the requirements of the problem.

@@

%\subsection{Importância de testes de hipóteses em regressão}

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

%\subsection{ANOVA e MANOVA}

In the case of traditional linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by \cite{anova_fisher}. According to \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each of the explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. And among the multivariate hypothesis tests already discussed in the literature, Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

%\subsection{Testes de comparações múltiplas}

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences are. For example, suppose there is a three-level categorical variable $X$ in the model: A, B, and C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used. Among the tests discussed in the literature are the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

%\subsection{Objetivo do trabalho}

However, for regression problems with multiple non-Gaussian responses, there are few alternatives for hypothesis testing. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is the development of hypothesis tests for McGLMs. We have the following specific objectives: to propose the use of the Wald test to carry out tests of general hypotheses on regression and dispersion parameters of McGLMs, also enabling the generation of tables of analysis of variance, multivariate analysis of variance and multiple comparisons tests for regression models with multiple non-Gaussian responses. Another objective is to evaluate the properties of the proposed tests based on simulation studies and to motivate the application potential of the discussed methodologies based on the application to the aforementioned datasets. In the problem, traditional modeling techniques would be difficult to apply, however it is a problem of possible analysis via McGLM and hypothesis tests can be used to evaluate the effect of the interaction between time and use of the probiotic on addiction and disorder of binge eating.

%\subsection{Organização do trabalho (seções)}

We present the data set in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model to the data and present the main results of the analysis of data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations.

@@

%\subsection{Contexto}

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

%\subsection{Desenho experimental e coleta de dados}

@@

%\subsection{Conjunto de dados}

For analysis purposes, the score that characterizes compulsion and the number of symptoms presented that characterize addiction were transformed to the unitary scale, considering that these are restricted variables. The purpose of the analysis is to assess the effect of moment and group on addiction and compulsion metrics. The dataset contains the following variables:

\begin{itemize}
  \item id: individual identifier variable.
  \item moment: moment identifier variable (T0, T1, T2).
  \item group: group identifier variable (placebo, probiotic).
  \item YFAS: proportion of symptoms that characterize addiction.
  \item BES: proporção de escore de compulsão.
\end{itemize}

%\subsection{Análise exploratória}

The graphical analysis presented in \autoref{fig1} shows, in (a) and (d) that both variables of interest present considerable asymmetry to the right. The boxplots of metrics as a function of group presented in (b) and (e) show significant differences between the placebo and probiotic groups for both responses. The boxplots of the metrics as a function of the evaluation moments, presented in (c) and (f), show that for both metrics the values were higher at T0, with a considerable reduction at T1. When comparing T1 and T2, for YFAS it seems that there is a slight increase in the last evaluation; for BES, T1 and T2 do not seem to differ.

@@

Still in an exploratory way, we can evaluate the behavior of addiction and compulsion metrics through the evaluation of descriptive measures by moment and by group, presented in \autoref{tab:tab1}. It is possible to verify the reduction of individuals over time, something common in prospective studies. Regarding the measurements, it is noted that both groups (placebo or probiotic) have higher means at T0 than at other times. Therefore, there is a clear reduction in metrics when compared to the preoperative period. When comparing the postoperative moments (T1 and T2) we verified that the YFAS measurements for the placebo and probiotic groups and BES in the placebo group showed, on average, an increase in the metrics at the last evaluation moment. On the other hand, the BES measures in the probiotic group showed a decrease.

@@

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ a matrix of response variables and $\boldsymbol{ M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$ is denoted by $\Sigma_r$, has dimension $NxN$. In addition, a correlation matrix $\Sigma_b$, of order $R \times R$, which describes the correlation between the response variables, is required.

The McGLMs \citep{Bonat16}are defined by:

@@

\noindent where the $g_r()$ functions are the traditional link functions; $\boldsymbol{X}_r$ denotes a design matrix $N \times k_r$; $\boldsymbol{\beta}_r$ denotes a vector $k_r \times 1$ of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\ boldsymbol{\Sigma}}_R^\top)$ is the generalized Kronecker product  \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

\noindent where $\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right) = diag(\vartheta(\boldsymbol{\mu}_r; p_r))$ denotes a diagonal matrix in which the entries are given by the variance function $\vartheta(\cdot; p_r)$ applied to the elements of the vector $\boldsymbol{\mu}_r$. Different choices of variance functions $\vartheta(\cdot; p_r)$ imply in different assumptions about the distribution of the response variable. We will mention 3 options of variance functions: power variance function, Poisson–Tweedie dispersion function and binomial variance function.

@@

The Poisson–Tweedie dispersion function \cite{Jorgensen15} is an option for responses that characterize counts. The dispersion function is given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. We thus have a rich class of models to deal with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

Finally, the binomial variance function, given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_{r1}}_r(1 - \mu_r)^{p_{r2}}$ is indicated when the response variable is binary, restricted to an interval, or when there is a number of successes in a number of trials.

It is possible to notice that the power parameter $p$ appears in all the variance functions discussed. This parameter is especially important because it is an index that distinguishes different probability distributions that are important in the modeling context and, for this reason, can be used as a tool for automatic selection of the probability distribution that best suits the problem.

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of \cite{Anderson73} and \cite{Pourahmadi00}, \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

\noindent where $h()$ is the covariance link function, $Z_{rd}$ with $d$ = 0,$\ldots$, D are matrices representing the covariance structure present in each response variable $r$ and $\boldsymbol{\tau_r}$ = $(\tau_{r0}, \ldots, \tau_{rD})$ is a vector $(D + 1) \times 1$ of dispersion parameters.

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by \cite{Demidenko13}.

In this way, the McGLMs configure a general framework for analysis via regression models for non-Gaussian data with multiple responses in which no assumptions are made regarding the independence of the observations. The class is defined by 3 functions (link, variance and covariance) in addition to a linear predictor and a matrix linear predictor for each response under analysis.

@@

McGLMs are fitted are fitted based on the estimating function approach described in detail by \cite{Bonat16} and \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

McGLMs' second-moment assumptions allow the division of parameters into two sets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^ {\top}$. So $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a vector $K \times 1$ of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top , \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a vector $Q \times 1$ of dispersion parameters. Also, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector of order $NR \times 1$ from the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^ \top)^\top$ denotes the stacked vector of order $NR \times 1$ of the expected value matrix $\boldsymbol{M}_{N \times R}$.

To estimate the regression parameters, the quasi-score function \cite{Liang86} is used, represented by

$$
\begin{aligned}
  \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
  \boldsymbol{\lambda}) = \boldsymbol{D}^\top
  \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
\end{aligned}
$$

>>>>

\title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.}

@@

\author[2]{Ligia de Oliveira Carlos}

\author[2]{Antônio Carlos Ligocki Campos}

\author[3]{Wagner Hugo Bonat}

@@

\address[2]{\orgdiv{Department of Surgical Clinic}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[3]{\orgdiv{Department of Statistics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

@@

Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

@@

\jnlcitation{\cname{%
\author{Freitas LAC}, 
\author{Carlos LO},
\author{Campos ACL} and 
\author{Bonat WH}} (\cyear{2022}), 
\ctitle{Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models}, \cjournal{Statistics in Medicine}}

@@

% Estudos clínicos

Clinical trials represent one of the most common procedures used to assess whether a new treatment is safe, effective or even better than a standard treatment. These are procedures widely used in medical research with the aim of evaluating interventions on a group of individuals. In general, the data generated by clinical trials are not trivial and, for analysis, there is a need for statistical methods suitable to deal with all the specifics of this type of study.

According to Meinert and Tonascia \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

% Problemas na área médica que envolvem múltiplas respostas

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: Kangovi et al \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; Song and Baicker \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; Thyregod et al \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); Schmitz et al \cite{schmitz2019effect} evaluated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; Bonat et al \cite{plastica} proposed a model to deal with multiple outcomes in repeated measures studies motivated by a problem that aimed to compare different aesthetic eyelid surgery techniques; Petterle et al \cite{petterle2021multivariate} proposed a multivariate regression model to handle multiple continuous bounded outcomes motivated by a body fat percentage data. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

% Introdução ao Dataset

Similarly to the aforementioned articles, we are interested in the analysis of multiple outcomes in the context of longitudinal data analysis. The study we shall describe was conducted with the objective of evaluating the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. This is a clinical trial with two response variables: a score that characterizes compulsion and the number of symptoms presented that characterize addiction. Both responses are limited counts and for analysis purposes have been transformed to unit range.

In this study, a set of subjects was divided into two groups: one of them received a placebo and the other received the treatment. In addition, individuals were evaluated at 3 different time points: the first evaluation was performed approximately 10 days before surgery; follow-up assessments were performed approximately three months and one year postoperatively.

% McGLM

It is important to note that this is a multivariate problem (contains two response variables), with non-Gaussian responses (both taking values in the unit range) and that the observations that make up the data set cannot be considered independent, considering that measurements taken on the same individual at different time points tend to be correlated. Therefore, it is a problem that standard modeling techniques would be difficult to apply. It is necessary to use some methodology that supports the design of the clinical trial adequately.

@@

% Importância de testes de hipóteses em regressão

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. Engle \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

% ANOVA e MANOVA

In the case of standard linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by Fisher and Mackenzie \cite{anova_fisher}. According to St, Wold et al \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and it is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. Among the multivariate hypothesis tests already discussed in the literature for MANOVAs, we have the Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

% Testes de comparações múltiplas

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences appear. For example, suppose there is a three-level categorical variable $X$ in the model that assumes the levels: A, B, or C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used, some of the tests discussed in the literature are: the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. Hsu \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. Bretz, Hothorn, and Westfall \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

% Objetivo do trabalho

However, for multiple response regression models, there are few alternatives for hypothesis testing. Therefore, as it is a flexible class of models with high power of application to practical problems, our general objective is the development of hypothesis tests for McGLMs. We have the following specific objectives: to propose the use of the Wald test to carry out tests of general hypotheses on regression and dispersion parameters of McGLMs, also enabling the generation of tables of analysis of variance, multivariate analysis of variance and multiple comparisons tests for regression models with multiple Gaussian and non-Gaussian responses. Another objective is to evaluate the properties of the proposed tests based on simulation studies and to motivate the potential application of the discussed methodologies based on the analysis of the aforementioned dataset. In the case study, standard modeling techniques would be difficult to apply, however it is a problem of possible analysis via McGLM and hypothesis tests can be used to evaluate the effect of the interaction between time and use of the probiotic on addiction and binge eating disorder.

% Organização do trabalho (seções)

We present the dataset in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations.

@@

% Contexto

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by Gormally et al \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by Gearhardt et al \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

% Desenho experimental e coleta de dados

@@

% Conjunto de dados

For analysis purposes, the score that characterizes compulsion and the number of symptoms presented that characterize addiction were transformed to the unitary scale, considering that these are counts variables restricted to a range. The purpose of the analysis is to assess the effect of moment and group on addiction and compulsion metrics. The dataset contains the following variables:

\begin{itemize}
  \item id: individual identifier variable.
  \item moment: moment identifier variable (T0, T1, T2).
  \item group: group identifier variable (placebo, probiotic).
  \item YFAS: proportion of symptoms that characterize addiction.
  \item BES: BES score transformed to unit scale.
\end{itemize}

% Análise exploratória

The graphical analysis presented in \autoref{fig1} shows, in (a) and (d) that both variables of interest present considerable asymmetry to the right. The boxplots of metrics as a function of group presented in (b) and (e) show little difference between the placebo and probiotic groups for both responses. The boxplots of the metrics as a function of the evaluation moments, presented in (c) and (f), show that for both metrics the values were higher at T0, with a considerable reduction at T1. When comparing T1 and T2, for YFAS it seems that there is a slight increase in the last evaluation; for BES, T1 and T2 do not seem to differ.

@@

Still in an exploratory way, we can evaluate the behavior of addiction and compulsion metrics through the evaluation of descriptive measures by moment and by group, presented in \autoref{tab:tab1}. It is possible to verify the reduction of individuals over time, something common in prospective studies. Regarding the measurements, it is noted that both groups (placebo or probiotic) have higher means at T0 than at other times. Therefore, there is a clear reduction in metrics when compared to the preoperative period. When comparing the postoperative moments (T1 and T2) we verified that the YFAS measurements for the placebo and probiotic groups showed, on average, an increase in the metrics at the last evaluation moment. The same could be seen for BES in the placebo group. On the other hand, the BES measures in the probiotic group showed a decrease.

@@

Consider $\boldsymbol{Y}_{N \times R} = \left \{ \boldsymbol{Y}_1, \dots, \boldsymbol{Y}_R \right \}$ be a matrix of response variables and $\boldsymbol{ M}_{N \times R} = \left \{ \boldsymbol{\mu}_1, \dots, \boldsymbol{\mu}_R \right \}$ be a matrix of expected values. The variance and covariance matrix for each response $r$, $r = 1,..., R$, is denoted by $\Sigma_r$ and has dimension $N \times N$. In addition, consider a correlation matrix $\Sigma_b$, of order $R \times R$, which describes the correlation between the response variables.

The McGLMs \citep{Bonat16} are defined by:

@@

\noindent where the $g_r()$ are standard link functions; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

\noindent where $\mathrm{V}\left(\boldsymbol{\mu}_r; p_r\right) = diag(\vartheta(\boldsymbol{\mu}_r; p_r))$ denotes a diagonal matrix in which the entries are given by the variance function $\vartheta(\cdot; p_r)$ applied to the elements of the vector $\boldsymbol{\mu}_r$. Different choices of variance functions $\vartheta(\cdot; p_r)$ imply different assumptions about the distribution of the response variable. We will mention 3 options of variance functions: power variance function, Poisson–Tweedie dispersion function and binomial variance function.

@@

The Poisson–Tweedie dispersion function \cite{Jorgensen15} is an option for responses that characterize counts. The dispersion function given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. Thus, we have a rich class of models to deal with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

Finally, the binomial variance function, given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_{r1}}_r(1 - \mu_r)^{p_{r2}}$ is indicated when the response variable is binary, bounded to an interval, or when there is a number of successes in a number of trials.

It is possible to notice that the power parameter $p$ appears in all the variance/dispersion functions discussed. This parameter is especially important because it is an index that distinguishes among different probability distributions that are important in the modeling context and, for this reason, can be used as a tool for automatic selection of the probability distribution that best fits to the data.

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of Anderson et al \cite{Anderson73} and Pourahmadi \cite{Pourahmadi00}, Bonat and Jørgensen \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

\noindent where $h()$ is the covariance link function, $Z_{rd}$ with $d$ = 0,$\ldots$, D are known matrices representing the covariance structure present in each response variable $r$ and $\boldsymbol{\tau_r}$ = $(\tau_{r0}, \ldots, \tau_{rD})$ is a vector $(D + 1) \times 1$ of dispersion parameters.

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by Pinheiro and Bates \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most well-known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by Demidenko \cite{Demidenko13}.

In this way, the McGLMs configure a general framework for analysis via regression models for Gaussian and non-Gaussian data with multiple responses in which no assumptions are made regarding the independence of the observations. The class is defined by 3 functions (link, variance and covariance) in addition to a linear predictor and a matrix linear predictor for each response under analysis.

@@

McGLMs are fitted based on the estimating function approach described in detail by Bonat and Jørgensen \cite{Bonat16} and Jørgensen and Knudsen \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

McGLMs' second-moment assumptions allow us to spplit the parameter vector into two sets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^ {\top}$. So $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a $K \times 1$ vector of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top , \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a $Q \times 1$ vector of dispersion parameters. Also, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector of order $NR \times 1$ from the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^ \top)^\top$ denotes the stacked vector of order $NR \times 1$ of the expected value matrix $\boldsymbol{M}_{N \times R}$.

To estimate the regression parameters, the following quasi-score function \cite{Liang86} is used, 
$$
\begin{aligned}
  \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
  \boldsymbol{\lambda}) = \boldsymbol{D}^\top
  \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
\end{aligned}
$$
